Follow-up Samples

### 2017-02-20 submission/manuscript_outline_20161024.Rmd

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.  We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure 3B].  For this model FIT measurement resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure 4B].  For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the sixth largest decrease in MDA [Figure 4B].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 4 & S4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 4 & S4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Interestingly, within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.  Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.

@@

***Lesion Model Creation:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  In brief, the model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Initial Sample Model Creation:***
We also investigated whether a model could be created that could identify pre (initial) and post (follow up) surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial surgery sample.

***Selection of Important OTUs:***
In order to assess which variables were most important to either model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or before sample models.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  A list of common OTUs were found for the two different models used.  Finally, both lesion and initial sample models initial positive probabilities were tested against the follow up positive probabilities.

@@

**Figure 3: Receiver Operating Characteristic Curve for Lesion and Initial Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set and what was used for subsequent classification.  A) Lesion model.  B) Before sample model

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S2: Summary of Important Variables in the Lesion Model** A) MDA of the most important variables in the lesion model.  The black point represents the median and the different colors are the different runs up to 100.  B) The total number of appearances of each variable in the 100 different lesion models.  The cutoff of 50% was used to assess importance.

**Figure S3: Summary of Important Variables in Before Sample Model** A) MDA of the most important variables in the lesion model.  The black point represents the median and the different colors are the different runs up to 100.  B) The total number of appearances of each variable in the 100 different lesion models.  The cutoff of 50% was used to assess importance.

**Figure S4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Full Variable Models** A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample in those with adenoma or Screen Relevant Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

**Figure S5: Thetayc Versus Time of Follow up Sample from Initial**

>>>>

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.  We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].  The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

Interestingly, within figure 4 the initial sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.  Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.

@@

***Lesion Model Creation:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.

In brief, the model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process. Next, in order to assess which variables were most important to the model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the model.  This reduced data set was then run through the mtry optimization again.  Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.

The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Initial Sample Model Creation:***
We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial to lesion removal sample.

@@

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.  Finally, a list of common OTUs were found for the two two different models used.

@@

**Figure 3: The Lesion Model.** A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.  B) Summary of Important Variables.  MDA of the most important variables in the lesion model.  The black point represents the median and the different colors are the different runs up to 100. C) Positive probability change from initial to follow up sample in those with carcinomaA) Lesion model.  D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN)

**Figure 4: The Initial Sample Model.**  A) ROC Curve: The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data and the blue line represents the model using 100% of the data set and what was used for subsequent classification.  B) Summary of Important Variables.  MDA of the most important variables in the initial sample model.  The black point represents the median and the different colors are the different runs up to 100.  C) Positive probability change from initial to follow up sample in those with carcinoma.  D) Positive probability change from initial to follow up sample if those with adenoma or Screen Relevant Neoplasia (SRN).

@@

**Figure S2: Thetayc Versus Time of Follow up Sample from Initial**

### 2017-02-20 submission/manuscript_outline_20161024.tex

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal control individuals. We investigated this by
creating a model to classify lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables {[}Figure 3B{]}. For this model FIT measurement
resulted in the largest decrease in MDA while the OTU with the largest
MDA was \emph{Lachnospiraceae} (Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we wanted to then build a separate model
specifically to be able to classify whether samples were initial or
follow up samples based on the bacterial community and FIT measurements.
The initial sample model had an AUC range of 0.641 to 0.805 after 100
iterations of 20 repeated 10-fold cross validations. The test set AUC
range for this model performed better then the training set AUCs. There
was a marked decrease in the ROC curve for the final model used when
compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
a total of 70 variables {[}Figure 4B{]}. For this model
\emph{Pseudomonas} (Otu000438) resulted in the largest decrease in MDA
while FIT measurement resulted in the sixth largest decrease in MDA
{[}Figure 4B{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 4 \& S4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 4
\& S4{]}. Providing additional information on the importance of
commensal bacteria was that there were a total of 23 OTUs in common to
both models and the vast majority belonged to regular residents of our
gut community {[}Table S3{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive ewithin our models. There are a number of reasons why this
may have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 4{]}.
It should be noted that our study does not argue against the importance
of these CRC associated bacteria in the pathogenesis of disease but
rather that they are not the main bacteria changing after surgical
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

@@

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals. In brief, the model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and OTUs that had near zero variance were
removed. This pre-processing was performed with the R package caret
(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
samples and making 100 different 80/20 (train/test) splits of the data
where normal and lesion were represented in the same proportion within
both the whole data set and the 80/20 split. Each of these splits were
then run through 20 repeated 10-fold cross validations to optimize the
mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 423 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) surgery samples. The main difference was that only the
67-person cohort was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyper-parameter was completed using the
same procedure as was used for the lesion model. Instead of classifying
samples as positive or negative of lesion this model classified samples
as positive or negative for being an initial surgery sample.

\textbf{\emph{Selection of Important OTUs:}} In order to assess which
variables were most important to either model we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different 80/20 split models and then
filtered this list to variables that were only present more than 50\% of
the time. This final collated list of variables was what was considered
the most important for the lesion or before sample models.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs that were used in
either built model, and investigating the relative abundance of specific
previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
the lesion model was then tested for accuracy in prediction and whether
it reduced the positive probability of lesion after surgery. The most
important OTUs for this were used to build a reduced model and it was
assessed for similarity to the original model. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial sample model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. A list of common OTUs were found
for the two different models used. Finally, both lesion and initial
sample models initial positive probabilities were tested against the
follow up positive probabilities.

@@

\textbf{Figure 3: Receiver Operating Characteristic Curve for Lesion and
Initial Sample Models.} The shaded areas represents the range of values
of a 100 different 80/20 splits of the test set data using either all
variables (grey) or reduced variable (red) models. The blue line
represents the reduced variable model using 100\% of the data set and
what was used for subsequent classification. A) Lesion model. B) Before
sample model

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample if those with adenoma or Screen
Relevant Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

@@

\textbf{Figure S2: Summary of Important Variables in the Lesion Model}
A) MDA of the most important variables in the lesion model. The black
point represents the median and the different colors are the different
runs up to 100. B) The total number of appearances of each variable in
the 100 different lesion models. The cutoff of 50\% was used to assess
importance.

\textbf{Figure S3: Summary of Important Variables in Before Sample
Model} A) MDA of the most important variables in the lesion model. The
black point represents the median and the different colors are the
different runs up to 100. B) The total number of appearances of each
variable in the 100 different lesion models. The cutoff of 50\% was used
to assess importance.

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Full Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample in those with adenoma or Screen
Relevant Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

\textbf{Figure S5: Thetayc Versus Time of Follow up Sample from Initial}

>>>>

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal control individuals. We investigated this by
creating a model to classify lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we wanted to then build a separate model
specifically to be able to classify whether samples were initial or
follow up samples based on the bacterial community and FIT measurements.
The initial sample model had an AUC range of 0.641 to 0.805 after 100
iterations of 20 repeated 10-fold cross validations. The test set AUC
range for this model performed better then the training set AUCs. There
was a marked decrease in the ROC curve for the final model used when
compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
a total of 70 variables that were used for this model {[}Figure 4B{]}.
The variable that resulted in the largest MDA was \emph{Pseudomonas}
(Otu000438) resulted in the largest decrease in MDA while FIT
measurement for fecal blood resulted in the sixth largest decrease in
MDA {[}Figure 4B{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 3
\& 4{]}. Providing additional information on the importance of commensal
bacteria was that there were a total of 23 OTUs in common to both models
and the vast majority belonged to regular residents of our gut community
{[}Table S3{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive ewithin our models. There are a number of reasons why this
may have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
surgical removal of lesion. In fact, it is possible that these CRC
associated bacteria are important in the transition from adenoma to
carcinoma and would be one explanation as to why in our data we not only
see high initial relative abundances in carcinoma and not adenoma
individuals but also large decreases in relative abundance in some of
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

@@

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals.

In brief, the model included data on FIT and the bacterial microbiome.
Non-binary data was checked for near zero variance and OTUs that had
near zero variance were removed. This pre-processing was performed with
the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
involved taking the samples and making 100 different 80/20 (train/test)
splits of the data where normal and lesion were represented in the same
proportion within both the whole data set and the 80/20 split. Each of
these splits were then run through 20 repeated 10-fold cross validations
to optimize the mtry hyper-parameter by maximizing the AUC (Area Under
the Curve of the Receiver Operator Characteristic). This resulting model
was then tested on the 20\% of the data that was originally held out
from this overall process. Next, in order to assess which variables were
most important to the model we counted the number of times a variable
was present in the top 10\% of mean decrease in accuracy (MDA) for each
of the 100 different 80/20 split models and then filtered this list to
variables that were only present more than 50\% of the time. This final
collated list of variables was what was considered the most important
for the model. This reduced data set was then run through the mtry
optimization again. Once the ideal mtry was found the entire 423 sample
set was used to create the final Random Forest model on which
classifications on the 67-person cohort was completed.

The default cutoff of 0.5 was used as the threshold to classify
individuals as positive or negative for lesion. The hyper-parameter,
mtry, defines the number of variables to investigate at each split
before a new division of the data is created with the Random Forest
model.

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) lesion removal. The main difference was that only the
67-person cohort was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyper-parameter was completed using the
same procedure as was used for the lesion model. Instead of classifying
samples as positive or negative of lesion this model classified samples
as positive or negative for being an initial to lesion removal sample.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs that were used in
either built model, and investigating the relative abundance of specific
previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
the lesion model was then tested for accuracy in prediction and whether
it reduced the positive probability of lesion after surgery. We then
used the initial sample model to assess whether it could classify
samples better then the lesion model and whether it could reduce the
positive probability of an initial sample in the follow up samples.
Finally, a list of common OTUs were found for the two two different
models used.

@@

\textbf{Figure 3: The Lesion Model.} A) ROC Curve: The shaded areas
represents the range of values of a 100 different 80/20 splits of the
test set data and the blue line represents the model using 100\% of the
data set and what was used for subsequent classification. B) Summary of
Important Variables. MDA of the most important variables in the lesion
model. The black point represents the median and the different colors
are the different runs up to 100. C) Positive probability change from
initial to follow up sample in those with carcinomaA) Lesion model. D)
Positive probability change from initial to follow up sample if those
with adenoma or Screen Relevant Neoplasia (SRN)

\textbf{Figure 4: The Initial Sample Model.} A) ROC Curve: The shaded
areas represents the range of values of a 100 different 80/20 splits of
the test set data and the blue line represents the model using 100\% of
the data set and what was used for subsequent classification. B) Summary
of Important Variables. MDA of the most important variables in the
initial sample model. The black point represents the median and the
different colors are the different runs up to 100. C) Positive
probability change from initial to follow up sample in those with
carcinoma. D) Positive probability change from initial to follow up
sample if those with adenoma or Screen Relevant Neoplasia (SRN).

@@

\textbf{Figure S2: Thetayc Versus Time of Follow up Sample from Initial}
